\subsection{Large Language Models}

LLMs such as GPT-3, GPT-4, and Claude transformed the landscape of NL2SQL
research. Early experiments with LLMs for NL2SQL tasks showed competitive
capabilities against PLM approaches \citep{DAIL-SQL}. \cite{T2SQL-LLM-Bench-3}
demonstrated that \textsc{Codex}, a GPT-3-based model, without fine-tuning,
could achieve competitive performance on \textsc{Spider}, outperforming
existing approaches that required extensive training.

\subsubsection{In-Context Learning}

ICL, sometimes referred to as few-shot learning, is an approach leveraging the
large context windows of LLMs by providing examples. The context windows of
frontier LLMs can reach up to hundred thousands of tokens, enabling researchers
to provide examples of accurate NL2SQL translation instead of applying
parameter updates. ICL in NL2SQL refers to the process of providing the LLM
with a small number of example pairs of NLQ to SQL mappings. This approach
provides flexibility, as database contents, previous answers and large datasets
can be dynamically utilized based on the question, rather than requiring
retraining.

Example selection strategies were shown to have a substantial impact on ICL
performance; \cite{DAIL-SQL} evaluated example selection methods like Random,
Question Similarity Selection, Masked Question Similarity Selection, and Query
Similarity Selection and proposed a strategy to select, organize and present
ICL examples to LLMs. DAIL-SQL utilizes both question and query similarity,
masking domain-specific words and prioritizing examples that exceed a
similarity threshold of $\tau$ \citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes
examples as question-SQL-pairs without the respective schema to improve token
efficiency. Using a Code Representation Prompt for question and schema encoding
yielded DAIL-SQL to achieve 86.6\% execution accuracy on \Spider. While GPT-4
achieves 72.3\% execution accuracy zero-shot on \textsc{Spider} \citep[Table 1,
p.~8]{DAIL-SQL}, few-shot learning shows to improve model performance. One-shot
boosts GPT-4's execution accuracy to 80.2\%, a 7.9\% increase, while five-shot
reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}. \textsc{XiYan-SQL, Chase-SQL
and Din-SQL} all utilize variations of ICL to achieve their results
\citep{XiYan, CHASE, DINSQL}.

\subsubsection{Self-Correction and Iterative Refinement}

\cite{DINSQL} proposed DIN-SQL as an approach to NLIDBs that rely on LLMs.
DIN-SQL decomposes complex queries into sub-parts and utilizes ICL and
self-correction during the generation phase. Compared to DAIL-SQL which relies
on example selection, DIN-SQL focuses on a refinement loop that allows the
model to self-correct errors, repairing schema linking, syntactic or semantic
errors by reviewing its output against the schema, user input and database
errors. DIN-SQL achieves 85.3\% execution accuracy on \textsc{Spider},
surpassed by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. \citeauthor{DINSQL}
found that generic self-correction which assumes that the query contains
errors, lowers execution accuracy by 4.2\% on \textsc{Spider} compared to
gentle self-correction, which assumes nothing about the validity of the query.
Smaller models perform better with generic and larger models with gentle
correction \citep{DINSQL}.

Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC
(Multi-Agent Guideline for In-Context Text-to-SQL) which automates
self-correction prompt engineering through a set of specialized agents
\citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a
feedback agent that collaboratively refine LLM instructions during the
refinement loop. MAGIC derives common failure patterns of the initial query
generation phase from training data, allowing it to spot common mistakes at
generation time. The autogenerated guidelines from MAGIC yield 85.6\% execution
accuracy on \textsc{Spider} (dev) which represented a 5.31\% improvement over
DIN-SQL's human written correction guidelines \citep{MAGIC}. While DIN-SQL and
MAGIC focus on automated self-correction in single-turn settings,
\cite{CoE-SQL} introduced Chain-of-Editions (CoE-SQL), which addresses
challenges of multi-turn conversational NLIDBs. Interactive information seeking
from the user has shown to be an effective way to drive accuracy and improve
user satisfaction \citep{NALIR}. Rather than approaching each query
independently, CoE-SQL recognizes that successive SQL queries in a
conversational context usually require only small and incremental modifications
of previous queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.

\subsubsection{Candidate Selection Frameworks}

Candidate selection strategies have shown performance improvements on
challenging benchmarks, acknowledging the difficulty of generating perfect SQL
in one attempt even with capable LLMs and self-correction mechanisms.

\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that
leverages multiple reasoning paths to generate query candidates. CHASE-SQL
harnesses three generation strategies: a divide-and-conquer approach which
breaks down complex NLQs into sub tasks, a Chain of Thought (COT) based
generation approach which inspects execution plans of SQL queries and a
schema-aware generation of synthetic examples for ICL \citep{CHASE}. For
candidate selection CHASE harnesses a fine-tuned LLM that can do binary
selection of candidates. CHASE-SQL yields 73\% execution accuracy on
\textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}. Conceptually similar
work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is
architected as multi-generation ensemble strategy with better schema
representation. \textsc{XiYan-SQL} integrated ICL alongside supervised
fine-tuning approaches to generate query candidates \citep{XiYan}. A key
contribution of \cite{XiYan} is their M-Schema representation of database
schemas, which improves the model's schema awareness and reduces schema linking
errors. The query generation stage utilizes both a fine-tuned SQL generation
model and ICL strategies. Following the query generation stage a
self-correction stage is utilized to correct common errors. Lastly a selection
model is used to choose the most accurate candidate \citep{XiYan}.
\textsc{XiYan-SQL} achieved 89.65\% execution accuracy on \textsc{Spider} and
73.34\% on \textsc{Bird} \citep{XiYan}.

Both CHASE and \textsc{XiYan-SQL} show that using distinct models for candidate
generation and candidate selection yield best-in-class execution accuracy,
combining the strengths of LLMs for language understanding with specialized
model architectures for candidate ranking.

\subsubsection{Retrieval-Augmented Generation}

Retrieval-Augmented Generation (RAG) has emerged as a paradigm for enhancing
LLMs and thus NL2SQL systems by integrating external knowledge retrieval with
the generative capabilities of LLMs. The most prevalent form of RAG in NL2SQL
is the encoding of the database schema into the LLM prompt. \cite{XiYan}
proposed the \textsc{M-Schema} format, a semi-structured text-based schema
description. \cite{DAIL-SQL} proposed a Code Representation Prompt encoding raw
SQL construction statements. \cite{XiYan} provided an ablation study for the
M-Schema which yielded questionable results. While the M-Schema format yielded
best results when \textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5
Sonnet, it performed worse than alternatives on DeepSeek and Gemini models
\citep{XiYan}.

\textsc{Resdsql} was one of the earlier approaches to explore subsetting
database schemas prior to encoding \citep{RESDSQL}. \cite{RetAug} introduced
\textsc{ASTReS} which dynamically retrieves database schemas and uses abstract
syntax trees (ASTs) to select optimal few-shot examples for ICL. By pruning the
ASTs down to the most relevant subset, \textsc{ASTReS} achieved 86.6\%
execution accuracy on \textsc{Spider} at time of publication, indicating that
schema subsetting is a sensible optimization mechanism. \textsc{ASTReS} was
combined with GRAPHIX-T5 to achieve this result \citep{RetAug}.

This technique becomes especially relevant for large enterprise schemas with
hundreds of tables and columns and recent benchmarks like \textsc{Spider2.0}.
\textsc{ASTReS} demonstrated that efficient schema retrieval enables smaller
models like GRAPHIX-T5 to achieve competitive performance against LLM-based
approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.

\subsubsection{Specialized LLMs and Fine-tuning}

While general-purpose LLMs demonstrated strong natural language capabilities,
researchers explored the potential of fine-tuned LLMs for NL2SQL tasks, which
offer a tradeoff between natural language understanding and SQL generation.

To resolve the prevalent dependency on proprietary APIs for inference,
\cite{CodeS} introduced \textsc{CodeS} in \citeyear{CodeS}, a series of
open-source language models dedicated for NL2SQL tasks with parameter
sizes ranging from 1B to 15B. \cite{CodeS} showed that \textsc{CodeS}-7B
achieves 85.4\% execution accuracy on \Spider, outperforming both
fine-tuning approaches like \textsc{Redsql} and GRAPHIX-T5-PICARD as well as
prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The
same tendency was observed on \textsc{Bird} with \textsc{CodeS}-15B achieving
60.37\%, compared to 57.41\% for DAIL-SQL+GPT-4 and 55.90\% for DIN-SQL+GPT-4
\citep{CodeS}. While more recent approaches like \textsc{XiYan-SQL} and CHASE
both outperform \textsc{CodeS}, CHASE relies on proprietary models and
\textsc{XiYan-SQL} doesn't disclose what base models were used.

\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which
introduced the next-generation \textsc{OmniSQL} models with 7B, 14B and 32B
sizes, trained using synthetic data generation. The models achieve execution
accuracy improvements on both \textsc{Spider} and \textsc{Bird}; the 7B model
reaches 88.9\% on \textsc{Spider} and 66.1\% on \textsc{Bird}, a 5\%+
improvement over comparable alternatives \citep{OmniSQL}. These results were
achieved without combining \textsc{OmniSQL} with advanced ICL, self-correction,
RAG or candidate-selection techniques, indicating that higher accuracy scores
are possible.

\subsubsection{Limitations and Challenges}

Despite the advancements of LLM-based NL2SQL systems, a set of limitations and
challenges emerged that prevent real-world adoption:

\begin{enumerate}
    \item \textbf{Hallucination} — LLM-based NL2SQL systems suffer from
        incorrect, but syntactically correct SQL queries. They produce subtle
        semantic errors like reversing the order of aggregations or using the
        wrong aggregation method are hard to catch. This is contrary to
        traditional approaches which fail visibly on unknown questions and
        databases. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
        Enterprise-Grade NL2SQL is still far from being resolved'' even with
        SOTA models, complex real-world database schemas and ambiguous queries
        remain challenging \citep{NL2SQLUnsolved, Spider2}.
    \item \textbf{Resource Requirements} — SOTA models require
        significant resources during inference, making them
        expensive to deploy at scale. While smaller LLMs exist, they
        typically show reduced performance on complex queries, creating a
        trade-off between accuracy and resource efficiency. Even fine-tuned
        LLMs like \textsc{OmniSQL} degrade in performance with shrinking
        parameter sizes, but maintain higher parameter efficiency over
        general purpose LLMs.
    \item \textbf{Data Privacy and Security Implications} — Using LLMs through
        proprietary APIs imposes data privacy and security threats as sensitive
        data may be transferred to respective model vendors. Recent open-source
        model developments like \textsc{CodeS} and \textsc{OmniSQL} mitigate
        this problem partially, but using local inference with LLMs increases
        hardware requirements.
    \item \textbf{Ambiguity Handling in Complex Scenarios} — LLMs continue to
        struggle with ambiguity resolution in NLQs, especially on complex
        database schemas. This becomes particularly prominent in enterprise
        databases where similar columns exist across multiple tables, or
        domain-specific terminology which might have multiple possible
        interpretations. Approaches like multi-path reasoning and candidate
        selection show improvements in execution accuracy, but often increase
        inference time and complexity
        \citep{CHASE, XiYan}.
    \item \textbf{Competitive PLM Approaches} — Approaches like
        \textsc{Resdsql} and GRAPHIX-T5 achieved competitive execution accuracy
        while using fewer parameters than LLMs \citep{RESDSQL}. Despite
        reducing inference-time requirements, these approaches required
        domain-specific fine-tuning which limited flexibility and introduces
        training costs. Using PLMs allows frontloading the computational effort
        and can thus reduce ongoing costs in compute-constrained environments.
        Yet, current research shifted towards LLMs due to their transferability
        and natural language capabilities.
\end{enumerate}
