\subsection{Research Gaps}

The literature shows that significant progress has been made in the development of
practically feasible NL2SQL systems. Despite these advancements, several critical
research gaps remain open which limit the practical deployment of NL2SQL
systems.

\subsubsection{Advanced Open-Source Approaches}

The research community primarily focused on developing the most performant
NL2SQL systems, paying less attention to fully open-source solutions. Thus,
two areas are under-researched:

\begin{enumerate}
    \item \textbf{Limited Exploration of Technique Synergies} — Research
        primarily focuses on techniques like candidate selection, RAG, and
        self-correction in isolation, using primarily proprietary LLMs.
        A gap remains in understanding how these techniques can be optimally
        combined, especially with open-source models.
    \item \textbf{Efficiency-Accuracy Tradeoffs} — Whilst recent research
        efforts focus primarily on achieving a SOTA execution accuracy metric,
        the relationship of computational requirements and performance gains
        remains unclear.
\end{enumerate}

\subsubsection{Deployment and Performance Gaps}

Despite impressive academic results, significant gaps remain in transitioning
NL2SQL systems from research environments to real-world deployments:

\begin{enumerate}
    \item \textbf{Database Integration} — While existing research has often
        focused on standalone systems, little attention has been given to the
        integration complexities between NL2SQL capabilities and DBMS.
    \item \textbf{Latency Considerations} — Most research papers
        evaluate models based on their accuracy metrics alone without factoring
        in the latency and throughput characteristics of their solutions,
        even though responsiveness is an important metric for user experience.
    \item \textbf{Privacy and Security} — While open-source models address some
        privacy concerns by enabling a local deployment of LLMs, research gaps
        remain regarding NL2SQL systems which respect database access controls
        and security policies.
\end{enumerate}

\subsubsection{Ambiguity Resolution and Semantic Accuracy}

Fundamental questions remain open regarding effective ambiguity
resolution. Current LLM-based systems struggle to identify queries containing
ambiguities that they can't confidently resolve. Current research and
benchmarks does not focus sufficiently on ambiguity detection and resolution.

\subsubsection{Evaluation and Benchmarking Gaps}

Two primary gaps remain in the evaluation and benchmarking infrastructure of
NL2SQL systems:

\begin{enumerate}
    \item \textbf{Performance Metrics} — Current benchmarks often don't capture
        latency or throughput of NL2SQL systems at all, allowing for solutions
        to achieve SOTA scores that require significantly more
        resources than their predecessors.
    \item \textbf{User Experience Metrics} — Contemporary benchmarks focus on
        execution accuracy without assessing potential user satisfaction,
        trust, and overall experience. Although execution accuracy is a big
        factor for the trustworthiness of a NL2SQL system, it is not the only
        component as indicated by \citeauthor*{NALIR} in \citeyear{NALIR}.
\end{enumerate}

\subsubsection{Thesis Placement}

The under researched aspects regarding the performance gap between proprietary
and open-source models as well as the integration, privacy and security
concerns mentioned above, motivate the implementation of \Natural. Specifically
this thesis researches the ability of schema-aware example selection for
in-context learning using graph-kernels to improve the semantic accuracy of
offline only NL2SQL systems. By providing a production-ready implementation of
a \Natural, this work further bridges the gaps between theoretical advancements
and their practical deployability, whilst exploring performance and portability
boundaries.
